---
title: Bear Braumoeller and Systemic Theory
output:
  md_document:
    variant: gfm
    preserve_yaml: TRUE
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../_posts") })
author: "Miles Williams"
date: "`r Sys.Date()`"
layout: post
categories: ["Methods", "Statistics"]
---

[Back to Blog](https://milesdwilliams15.github.io/blog/)

```{r setup, include=FALSE, cache=FALSE}
base_dir <- "~/My Website/milesdwilliams15.github.io/" # i.e. where the jekyll blog is on the hard drive.
base_url <- "/" # keep as is
fig_path <- "assets/images/2023-05-23/" # customize to heart's content, I 'spose.

knitr::opts_knit$set(base.dir = base_dir, base.url = base_url)
knitr::opts_chunk$set(fig.path = fig_path,
                      cache.path = '../cache/',
                      message=FALSE, warning=FALSE,
                      cache = F,
                      dpi = 700) 
library(tidyverse)
library(coolorrr)
set_theme()
```

The untimely passing of Dr. Bear Braumoeller still fills me with disbelief. As the [many kind words](https://twitter.com/osupolisci/status/1653857431188029440?lang=en) of individuals on Twitter attest, the OSU professor was not only a giant in the field of security studies but also a kind and supportive colleague and friend to many. 

On only one occasion did I have a chance to meet Bear in person. When I was a grad student at UIUC, I gave Bear a ride to the political science department from his hotel back in fall of 2019 during a visit. He was there to talk about his [latest book](https://global.oup.com/academic/product/only-the-dead-9780190849535?cc=us&lang=en&), an important and accessible read making a data-driven argument about the persistence of international war in the modern era. It was a powerful rebuttal to some claims made by other prominent scholars espousing the so-called "decline of war thesis," which holds that war is going out of fashion. Bear's book relies on a distinct mix of advanced statics and diplomatic history to convince you otherwise. 

I quickly devoured the book before Bear arrived at UIUC and hoped to talk with him about it. Unfortunately, as a star-struck third-year Ph.D. student at the time, I choked during the car ride and could muster little more than small talk.

It's hard meeting your heroes, and Bear was one of mine. His first book, [*The Great Powers and the International System*](https://assets.cambridge.org/97811070/05419/frontmatter/9781107005419_frontmatter.pdf), is just as timeless a read as his second book. When I first read it as a masters student back in 2016, it solidified my interest in studying International Relations. I knew then that I wanted to make a career out of studying big questions in international politics. Bear would never know it, but he's the reason I knew I wanted to get a Ph.D. in political science and pursue an academic career.

Years later, I'm (thankfully!) an employed academic at Denison University. And even more, given its location just outside Columbus, OH, I now live within only a few square miles of Bear's stomping grounds. After using his second book for one of the classes I taught in the Fall of 2022, I decided to reach out to him via email. I told him not only that I had a fun time using the book in class, but that I also created a [web application](https://milesdwilliams15.shinyapps.io/correlates-of-order/) based on the data he used in the book so my students could explore the data. I also took the opportunity to solicit some of his feedback on a [research note](https://github.com/milesdwilliams15/The-power-law-of-war/blob/main/04_report/01_combined_manuscript.pdf) his book inspired me to write dealing with the some of the technical details of statistically measuring trends in the intensity of war.

Bear's response was generous, and he expressed excitement about the app I created and provided me with helpful feedback on my research note, offering to take a second look if I chose to continue with project. Since he was in Oslo working on research of his own, I didn't have a chance to meet with him in person in Columbus, but hoped I might be able to touch base with him when he returned.

Sadly, this was not to be as Bear passed away in May of 2023. If my short correspondence with Bear was enough to make his passing a shock, the effect on the broader community of scholars in his network was exponentially greater---not to mention his family.

I'm in no position to offer a eulogy. I'm far too new and peripheral a member of Bear's network. Instead, in my own weird way, I want to honor his memory by writing post on how to program Bear's famous (but I feel under-appreciated) model of systemic politics. This model forms the foundation of his first book as well as [an article](https://www.cambridge.org/core/journals/american-political-science-review/article/abs/systemic-politics-and-the-origins-of-great-power-conflict/22394D8AB869CDF8F73735EBCF3A37E5) he published a few years prior in *APSR*.

As my friend Ekrim Baser [tweeted](https://twitter.com/ekrembaser/status/1654859887191502848) after Bear's passing, Bear's systemic approach wasn't given the attention it deserved. "IR *desperately* needs more general equilibrium thinking. We should be rewarding the precious few people who are crazy enough to push us in that direction," Ekrim further noted. 

Bear Braumoeller was one of these people, and I dared to be in my [dissertation](https://docs.google.com/viewer?url=https://github.com/milesdwilliams15/Dissertation/raw/main/dissertation_final.pdf) in the context of theorizing about the absence of widespread collusion in international development. I know from experience that you have to be a little crazy to engage in "general equilibrium thinking," but the payoffs for it are exponential. I also know that one of the main barriers to this approach is tractability. 

As inaccessible Bear's systemic theory was to me upon my first reading, it became far clearer when I started to get my hands dirty with it. By "getting my hands dirty," I mean programming. As with most advanced concepts, I personally find that I don't truly understand them until I can program them. That's just what I did with Bear's model of systemic politics, and detailing how I did this is what I'd like to do with the remainder of this post.

So, without further ado...

# What is Systemic Politics?

Braumoeller's model of systemic politics was his answer to the ["agent-structure problem"](https://www.jstor.org/stable/2706749?seq=1) in international politics. Do agents act upon their environment, or does their environment determine their behavior? The answer of course is that agents both shape, and are shaped by, the structures around them. Recognition of this fundamental truth is what makes Braumoeller's theory truly *systemic*, namely, that it characterizes the dynamic feedback loop between structure and agents.

Because it reflects "big picture" thinking, the theory's predictions are not so fine-grained as to be able to forecast the foreign policy activities of states or the occurrence and outcomes of wars. Rather, the theory offers predictions about how active states will be in international politics (from isolationist to highly internationalist) and how their activities will shape the distribution of different things states care about, like power or the spread of ideology.

This is not to say that systemic politics offers nothing of value to questions about the specific foreign policy choices of states. It can be easily combined with lower-level theories to explain the specific forms that a state's activity might take, as Braumoeller does in [other work](https://www.cambridge.org/core/journals/american-political-science-review/article/abs/systemic-politics-and-the-origins-of-great-power-conflict/22394D8AB869CDF8F73735EBCF3A37E5) examining militarized interstate disputes. Nonetheless, the generic framework is not tooled specifically to deal with the day-to-day choices of leaders or the outcomes of those choices. It is more sweeping than that, dealing instead with the process by which the international system comes to settle into a general equilibrium between the activity levels of states and the structures they inhabit.

# The Mathematical Model

The brilliance of Braumoeller was his ability to capture the interplay of agents and structure in the form of a mathematical model. While this model describes the activities of agents and the structures around them, it is not game theoretic. It is entirely devoid of any specification of actor utility functions. In this way, it is primarily *descriptive*, which is not the same as saying it is devoid of theory or mechanisms. 

The model itself is comprised of a system of differential equations relating the activities of an arbitrary number of countries to the state of the world along an arbitrary number of dimensions in the international system. For a country $c \in {1, ..., c, ..., N}$, its level of activity $a_{c} \geqslant 0$ at time $t+1$ is given as:

$a_{c(t+1)} = \Sigma_{d=1}^M\omega_{cd}[ v_c(c_{cd}) - s_{d(t)} ]^2$

In the above, state $c$'s activity level at a point in time is a function of its level of dissatisfaction between the state of the system ($s_{d}$) along each dimension $d \in {1, ..., d, ..., M}$ (these might be the distribution of power, ideology, etc.) at time $t$ and its ideal point for a given dimension ($v_c(c_{cd})$). The value $c_{cd}$ is a frequency distribution of the attitudes of individuals within a country with respect to a given dimension of the international structure, and $v_c(\cdot)$ is a preference aggregation function that maps the attitudes of individuals to a country-level preference about the the state of the system in dimension $d$. The weight placed on a given dimension is determined by the state's worldview $\omega_{cd} \geqslant 0$.

We can see from the above that the extent to which a country is internationalist versus isolationist is proportional to (1) the difference between the state of the world and a country's ideal state of the world and (2) the weight the state attaches to different dimensions of the state of th world.

The change produced by the collective activities of $N$ countries in dimension $d$ is then defined as:

$\Delta s_{d(t)} = \Sigma_{c = 1}^N \pi_c\omega_{cd}a_{c(t)} [ v_c(c_{cd}) - s_{d(t)}]$

Or, if you like:

$s_{d(t+1)} = \Sigma_{c = 1}^N \pi_c\omega_{cd}a_{c(t)} [ v_c(c_{cd}) - s_{d(t)}] + s_{d(t)}$

In the above, the difference between a country's ideal point and the state of the system determines the direction that a country's activity pushes the state of the system along a given dimension. The relative force behind this push is determined by the state's dimension-specific dissatisfaction, overall level of activity (the more internationalist, the stronger the impact), the strength of the state's worldview, and a new factor ($\pi_c$) which denotes the state's realized capabilities scaled to $0 \leqslant \pi_c \leqslant 1$.

These equations, depending on the starting values of the model's exogenous parameters, allow for a dynamic feedback loop between the state of the system and the level of activity of countries until the system eventually settles into an equilibrium where activity levels and the state of the world remain constant.

# Programming and Simulating the International System

This mathematical model lends itself well to simulation. It can create some headaches when it comes to programming it, however. So I'll walk through my approach slowly, explaining the code along the way. I've done the programming in R, restricting myself to `{baseR}` as much as possible, saving the `{tidyverse}` for visualization.

First, we need to create a function for activity given country world views, ideal points, and the current state of the system. The below function will accept a matrix `w`, a matrix `v`, and a vector `s` that provide the relevant information. This function is generalized to any arbitrary number of structural dimensions and countries. For `w` and `v`, information in columns will be for each country in the international system, and rows will denote dimensions for the state of the world. `s` will just be a vector denoting the starting values for the state of the system for each of $M$ arbitrary dimensions.

```{r}

## Create function for agent activity:
activity <- function(w,v,s){
  a <- 0
  for(i in 1:ncol(w)) {
    a[[i]] <- sum(w[, i] * (v[, i] - s)^2)
  }
  a
}
```

Say we have the following values for a pair of countries and one dimension characterizing the state of the world. If we test our activity function with the below inputs, we should see that country A is isolationist while B is internationalist. The reason is that, in the below code, the state of the world along the one dimension of relevance to our countries is equal to A's ideal point, but not B's.

```{r}
w <- cbind(c(0.5), c(0.5))
v <- cbind(c(0.5), c(1))
s <- 0.5
activity(w, v, s)
```

Next, we need a function that tells us how the structure of the system changes given the activity of countries. The below function takes the same inputs as before (`w`, `v`, and `s`) and adds a vector `a` for country activity levels and a vector `p` for county's relative realized capabilities.

```{r}

## Create function for structure:
structure <- function(a,p,w,v,s){
  d.s <- 0
  for(d in 1:length(s)) {
    d.s[d] <- sum(p * w[d, ] * a * (v[d, ] - s[d]))
  }
  d.s + s
}
```

If we run this function using the output from the activity function and the other values specified above with the added specification of relative capabilities, the function returns a new state of the world for time $t + 1$.

```{r}
p <- c(0.5, 0.5)
a <- activity(w, v, s)
structure(a, p, w, v, s)
```

Finally, we need a function that will let activity and structure play out over time. The below function lets the system run for a default of 20 time periods. It returns a list of objects: a data frame of country activity levels over time and a data frame of the state of the system over time, respectively.

```{r}

## Create function to let the system evolve
system <- function(p,w,v,s,periods=20){
  s.start <- s
  a.out <- list()
  s.out <- list()
  for(t in 1:periods) {
    a <- activity(w=w,v=v,s=s)
    s <- structure(a=a,p=p,w=w,v=v,s=s)
    a.out[[t]] <- data.frame(
      a = a,
      t = t,
      state = LETTERS[1:ncol(w)]
    )
    s.out[[t]] <- data.frame(
      s = s,
      t = t+1,
      dimension = as.factor(1:length(s))
    )
  }
  s.out.start <- data.frame(
    s = s.start,
    t = 1,
    dimension = as.factor(1:length(s))
  )
  s.out <- rbind(
    s.out.start,
    do.call(rbind,s.out)
  )
  list(
    activity = do.call(rbind,a.out),
    structure = s.out
  )
}
```

Given the above values, we can check what this looks like:

```{r}
system(p, w, v, s, periods = 5)
```

Finally, to make plotting more convenient, the below function will take the output from `system()` and show how activity and the state of the system change over time. (Here's the bit where I deviate from only using `{baseR}`.)

```{r}
## Create a function to plot how the system changes
library(tidyverse)
library(patchwork)
sys_plot <- function(sys.out,
                    xlab='time',
                    ylab=c('activity','structure'),
                    str.names,
                    act.names) {
  p1 <- sys.out$activity %>%
    ggplot() +
    aes(
      t,a,linetype=state
    ) +
    geom_line() +
    labs(
      x = xlab,
      y = ylab[1],
      linetype = "Country"
    ) +
    scale_x_continuous(
      breaks = NULL
    ) +
    scale_y_continuous(
      breaks = NULL
    ) +
    theme_classic() +
    theme(
      legend.position = 'top',
      legend.title = element_text(face=4)
    )
  p2 <- sys.out$structure %>%
    ggplot() +
    aes(
      t,s,linetype=dimension
    ) +
    geom_line() +
    scale_x_continuous(
      breaks = NULL
    ) +
    scale_y_continuous(
      breaks = NULL
    ) +
    labs(
      x = xlab,
      y = ylab[2],
      linetype = "Balance of..."
    ) +
    theme_classic() +
    theme(
      legend.position = "top",
      legend.title = element_text(face=4),
      axis.text.y = element_text(angle = 90, hjust = .5)
    )
  if(!missing(act.names)) p1 +
    scale_linetype(
      breaks = unique(sys.out$activity$state),
      labels = act.names
    ) -> p1
  if(!missing(str.names)) p2 + 
    scale_linetype(
      breaks = unique(sys.out$structure$dimension),
      labels = str.names
    ) -> p2
  p1 + p2
}

```

Let's take the above example, and show how it plays out. Note how little code I need to write (now that I've defined all the above functions) to run and visualize the simulation.

```{r}
sys_out <- system(p, w, v, s, periods = 100)
sys_plot(sys_out)
```

Given the starting parameters, the system moves closer to a balance between the preferences of country A and country B over time. As this happens, the former moves from being isolationist to more internationalist while the latter remains internationalist but with declining magnitude.

We can consider a more complex example, as well. Say we have a system with three countries and two dimensions characterizing the state of the world. These functions can easily generalize to this more complicated scenario.

```{r}

## Define the starting parameters
w <- cbind(w1 = c(.5,.5), w2 = c(.7,.3), w3 = c(0.9, 0.1))
v <- cbind(i1 = c(1,1), i2 = c(0,0), i3 = c(0.3, 0.3))
s <- c(.3,.3)
p <- c(.7,.3, .1)
p <- p / sum(p)

## Let the system run
sys_out <- system(p=p,w=w,v=v,s=s)

## Plot the system over time
sys_plot(sys_out)

```

It's that simple!

[Back to Blog](https://milesdwilliams15.github.io/blog/)